Elaboration with First-Class Implicit Function Types

### 2020-05-10 paper.tex
revs

\subsection{Core syntax}

@@

The universe $\U$ is Russell-style, and we have the type-in-type rule.
This causes our core syntax to be non-total, and our elaboration algorithm to be
possibly non-terminating. We use type-in-type to simplify presentation, since
consistent universe setups are orthogonal to the focus of this work.

@@

For $\einfer{x}{\Theta}{\Gamma}$, we look up the type of $x$ in $\Gamma$. In the
case of $\U$, we always succeed and infer $\U$ as type. In the cases for
function types, we check that the domains and codomains have type $\U$.  For
$\lambda$-s, we create a fresh meta for the domain type (since our surface
$\lambda$-s are not annotated), and infer types for the bodies.

@@

@@

We do not present any \emph{completeness} result for elaboration in this
paper. For an example of what this would entail, in \cite{dunfield2013complete}
completeness means that whenever there is a way to fill in missing details in
the surface syntax, algorithmic typechecking \emph{always} finds it. In
ibid.\ this means figuring out domain types for $\lambda$-s and inserting all
implicit applications. However, our elaborator targets a far stronger theory,
and it is beyond our reach to succinctly characterize which annotations are
inferable in the surface language, and we do not know of any prior work which
accomplishes this for a comparably strong elaborator. Our experience in Agda is
that it is not tractable in general to figure out which arguments are inferable,
by looking at function types, and often we have to run elaboration to see what
works.

@@

Telescopes and records are derivable from natural numbers, the unit type and
$\Sigma$-types. We write $\top$ for the unit type with $\tt : \top$. We use
Agda-like pattern matching notation in the following.  First, we define
length-indexed telescopes.
\begin{alignat*}{3}
  & \rlap{$\Tel' : \Nat \to \U$}\\
  & \Tel'\,\zero      && :\equiv \top\\
  & \Tel'\,(\suc\,n)  && :\equiv \Sigma(A : \U).\,(A \to \Tel'\,n)
\end{alignat*}
Then, we have $\Tel :\equiv \Sigma(n : \Nat).\,(\Tel'\,n)$, and define records:
\begin{alignat*}{3}
  & \rlap{$\Rec : \Tel \to \U$}\\
  & \Rec\,(\zero,\,\_)        &&:\equiv \top\\
  & \Rec\,(\suc\,n,\,(A,\,B)) &&:\equiv \Sigma(a : A).\,(\Rec\,(n,\,B\,a))
\end{alignat*}
From the above, $\epsilon$, $\blank\TCons\blank$, $\blank::\blank$ and ${[}{]}$
are evident, and all expected equalities hold definitionally. Derivability is
good news because we inherit nice properties of the type theory which only
contains the base type formers. For instance, if we have a consistent universe
setup\footnote{Recall that we currently assume type-in-type, which causes consistency and normalization to fail.}, we inherit consistency, canonicity and decidability of conversion. We
currently use native telescopes instead of $\Nat$, $\top$ and $\Sigma$ because
in unification and elaboration it is convenient that we are able to restrict
some types to records types.

@@

Hence, we can derive a somewhat weaker version of curried functions, with
propositional $\beta$ and $\eta$. From this, we still get consistency and
canonicity very cheaply. This is because models proving consistency or
canonicity for the base theory with $\Nat$, $\top$ and $\Sigma$ usually support
equality reflection. For consistency, standard set-theoretical models and models
in extensional type theory have this property, for canonicity, glued models
(e.g.\ as in \cite{kaposi2019gluing} and \cite{sterling2019algebraic}) also have
this property.

@@

Now, let us assume that $B'$ is not an implicit function type, curried function
type or meta-headed. Then, we have the following case, where we solve a
telescope domain to be empty.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,B' :\equiv \edo\\
& \quad (\Theta_1,\,\theta_1) \leftarrow \unify_{\Theta_0|\Gamma}\,A\,\epsilon\\
& \quad \unify_{\Theta_1|\Gamma[\theta_1]}\,(B[\theta_1][x\mapsto[]])\,(B'[\theta_1])
\end{alignat*}
Again, we also have the symmetric case. For $\lambda\,\{x : \ol{A}\}.\,t$ and
$t\,\{u : \ol{A}\}$, unification is structural, and other cases remain the same
as in the basic elaborator of Section \ref{sec:bidirectional_elaboration}.

>>>>

\subsection{Core syntax}
\label{sec:core}

@@

The universe $\U$ is Russell-style, and we have the type-in-type rule.  This
causes our core syntax to be non-total, and our elaboration algorithm to be
possibly non-terminating. Consistent universe setups are straightforward to
devise, but they are orthogonal to the focus of this work, and we choose to
avoid needless noise.

@@

For $\einfer{x}{\Theta}{\Gamma}$, we look up the type of $x$ in $\Gamma$. In the
case of $\U$, we always succeed and infer $\U$ as type. In the cases for
function types, we check that the domains and codomains have type $\U$.

@@

For $\lambda$-s, we create a fresh meta for the domain type (since our surface
$\lambda$-s are not annotated), and infer types for the bodies. Note the eager
insertion in $\einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A})$. This
causes, somewhat unexpectedly, the following example to fail:
\begin{alignat*}{3}
  & \slet\,f : \_ = \lambda\,A\,\{B\}.\,A\,\sin\,\\
  & \slet\,g : \U \to \{B : \U\}\to U = f\,\sin\,...
\end{alignat*}
The issue here is that we immediately apply the elaborated $\lambda\{B\}.\,A$
expression to a fresh meta, which causes the implicit function type to disappear
from the result, and preventing types to unify on the following line. We could
repair this by blocking insertion for implicit $\lambda$-s, but we choose to
keep the algorithm as simple as possible, and in any case overeager insertion
can be fixed by using type annotations to stay in checking mode.

@@

We do not present any \emph{completeness} result for elaboration in this
paper. For an example of what this would entail, in \cite{dunfield2013complete}
completeness means that whenever there is a way to fill in missing details in
the surface syntax, algorithmic typechecking \emph{always} finds it. In
ibid.\ this means figuring out domain types for $\lambda$-s and inserting all
implicit applications. However, we target a far stronger theory, and to our
knowledge no prior work has attempted to characterize inferable terms or show
completeness for a comparably strong elaborator. Our experience in Agda is that
it is not tractable in general to figure out which arguments are inferable, by
looking at function types, and often we have to run elaboration to see what
works. This does not imply that type inference with dependent types must
necessarily be an unwieldy affair. While we lose completeness, we gain the
ability to infer more parts of programs by following richer type dependencies,
as well as more power to refine and synthesize terms when types are known
beforehand.

@@

\subsubsection{On metatheoretical properties of the core syntax}
We naturally want to make sure that the extended core syntax is still sensible.
In relation to type theories, the following three properties are commonly
considered.
\begin{itemize}
  \item A theory is \emph{consistent} if not every closed type has a closed
    inhabitant. An inconsistent theory cannot be used as a \emph{logic}, from
    the propositions-as-types perspective \cite[Chapter~1]{hottbook}, because
    there every proposition is provable, so proofs do not distinguish truth from
    falsehood.
  \item A theory has the \emph{canonicity} property, if every closed term is
    definitionally equal to a canonical term. For example, every closed term
    with a natural number type is a numeral, every $\mathsf{Bool}$ is $\true$ or
    $\false$ and so on. A canonical theory is useful as a \emph{total
      programming language}; if canonicity fails, then some programs compute
    to stuck terms which cannot be viewed as values. In non-total languages,
    progress and type preservation constitute a weaker analogue of canonicity.
  \item A theory has \emph{decidable conversion}, if definitional equality is
    decidable. Undecidable conversion usually implies undecidable type checking
    for type theories. Often, decidable conversion follows from a normalization
    procedure for open terms: we first normalize, then decide equality of normal
    forms.
\end{itemize}

The properties are listed roughly in the order of increasing
strength. Canonicity often implies consistency (because empty types have no
canonical terms), and normalization implies canonicity. Now, our current core
theory has \emph{none} of the above properties, because the $\U : \U$ rule
precludes them. However, as we mentioned in Section \ref{sec:core}, this is
easily fixed by stratified universes.

In any case, a \emph{reduction} of new features to old features is a common and
convenient strategy to reduce metatheoretical properties of extended theories to
that of smaller or more conventional theories. In the simplest case, new
features are definable from old features. Here, we can either assume that new
features are merely definitional shorthands, or give an evident injective
translation from the new theory to the old one. Both imply that the desired
properties transfer from the old theory to the new one.

Telescopes and records are derivable from natural numbers, the unit type and
$\Sigma$-types. We write $\top$ for the unit type, and we have $\tt : \top$. We
use Agda-like pattern matching notation in the following.  First, we define
length-indexed telescopes.
\begin{alignat*}{3}
  & \rlap{$\Tel' : \Nat \to \U$}\\
  & \Tel'\,\zero      && :\equiv \top\\
  & \Tel'\,(\suc\,n)  && :\equiv \Sigma(A : \U).\,(A \to \Tel'\,n)
\end{alignat*}
Then, we have $\Tel :\equiv \Sigma(n : \Nat).\,(\Tel'\,n)$, and define records:
\begin{alignat*}{3}
  & \rlap{$\Rec : \Tel \to \U$}\\
  & \Rec\,(\zero,\,\_)        &&:\equiv \top\\
  & \Rec\,(\suc\,n,\,(A,\,B)) &&:\equiv \Sigma(a : A).\,(\Rec\,(n,\,B\,a))
\end{alignat*}
From the above, $\epsilon$, $\blank\TCons\blank$, $\blank::\blank$ and ${[}{]}$
are evident, and all expected equalities hold definitionally. Hence, it does not
matter whether telescopes are primitive or derived. In the rest of the paper we never
need to look inside the representation of telescopes, and we only use the ``interface''
from Figure \ref{fig:telescopes}.

@@

Hence, we can derive a somewhat weaker version of curried functions, with
propositional $\beta$ and $\eta$. From this, we still get consistency and
canonicity very cheaply. This is because models proving consistency or
canonicity for the base theory with $\Nat$, $\top$ and $\Sigma$ usually support
equality reflection, i.e.\ that propositional equality implies definitional
equality. For consistency, standard set-theoretical models and models in
extensional type theory have this property, for canonicity, glued models
(e.g.\ as in \cite{kaposi2019gluing} and \cite{sterling2019algebraic}) also have
this property.

@@

Now, let us assume that $B'$ is not an implicit function type, curried function
type or meta-headed. Then, we have the following case, where we solve a
telescope domain to be empty.
\begin{alignat*}{3}
& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,B' :\equiv \edo\\
& \quad (\Theta_1,\,\theta_1) \leftarrow \unify_{\Theta_0|\Gamma}\,A\,\epsilon\\
& \quad \unify_{\Theta_1|\Gamma[\theta_1]}\,(B[\theta_1][x\mapsto[]])\,(B'[\theta_1])
\end{alignat*}
Again, we also have the symmetric case. For telescopes, $\Rec$, inhabitants of
$\Rec$, curried abstraction and application, unification is structural, and
other cases remain the same as in the basic elaborator of Section
\ref{sec:bidirectional_elaboration}.
